---
title: "Join Lab"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---


# Joins and `dplyr` practice

Load the `"nycflights13"` package.  You'll also need to load `dplyr` or `tidyverse`, if you didn't above.

```{r}
library(tidyverse)
#install.packages("nycflights13")
library(nycflights13)
my_flights <- flights # store my own version
my_flights <- na.omit(my_flights) ## this dataset has data on cancelled flights too, 
                                  ## so we 'll remove those for now
library(dplyr)
```


Create a dataframe of the average arrival delays for each _destination_

```{r}
glimpse(my_flights)
avg_arrival_data <- my_flights %>%
                    group_by(dest) %>%
                    summarize("Avg Arrival Delays" = mean(arr_delay))
avg_arrival_data                 
```

Then use `left_join()` to join on the "airports" dataframe, which has the airport information.

```{r}
glimpse(airports)
airports_avg_arrival <- avg_arrival_data %>%
                        left_join(airports, by = c("dest" = "faa")) #in help notes dest = faa

airports_avg_arrival
```


Note here: the column names don't match anymore! 

Which airport had the largest average arrival delay?

```{r}
max_avg_arr <- airports_avg_arrival %>%
              arrange(desc(`Avg Arrival Delays`)) %>%
              slice(1)
              
              
max_avg_arr
```


What is the average departure delay for each airline? To answer this with something more than just the FAA code of the airline name, you need to create a dataframe of the average departure delay for each _airline_, then use `left_join()` to join on the "airlines" dataframe


```{r}
avg_dep_per_airline <- my_flights %>%
                      group_by(carrier) %>%
                      summarize("Average Departure Delay" = mean(dep_delay, na.rm = TRUE)) %>%
                      left_join(airlines, by = "carrier")
avg_dep_per_airline
                      

```

Which airline had the smallest average departure delay? 



```{r}
smallest_avg_dep_delay <- airports_avg_arrival %>%
              arrange(`Avg Arrival Delays`) %>%
              slice(1)

smallest_avg_dep_delay
```





Examples from R4DS:    

```{r}

top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest

```

Now you want to find each flight that went to one of those destinations. You could, of course, filter. But! Now you can semi-join: 

```{r}
flights %>% 
  semi_join(top_dest)
```


What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?
Perform the two joins below and consider what the results are. 
```{r}
anti_join(flights, airports, by = c("dest" = "faa")) 
#tells me all the flights that dont have airports in the airport dataset.

anti_join(airports, flights, by = c("faa" = "dest"))
#tells me all the airports that dont have any flights arriving.
```

