---
title: "Joins"
author: "K. Shoemaker"
date: "DATA 2401"
output:
  slidy_presentation: default
  pdf_document: default
  html_document: default
---

```{r, echo = F, message = F}
library(tidyverse, warn.conflicts = F)
```

# Joins

Sometimes you store data in different dataframes. Consider: you're a charity, and you have one data frame for donations, and one for email and phone number of your donors.

- This helps with data storage: If you store every donation with the email and phone number, you're wasting space. 

- This helps with data updating: If you need to update a donor's phone number, you only have to do it in one spot, not on every row of donations they've done. 

*Those of you who have worked in SQL, this type of thinking should look familiar.* 

--- 

So, how do we do this? 

You join data through data that is in both datasets: you need a way to tell the program how to join them. 

You identify columns that are in both tables, then use these columns to "match" the corresponding rows. 

---

So, suppose we have this table of donations: 

|Name| Amount | Date |
| --- | --- | --- 
| Sam | 100 | 2/12/2019 |
| Mike | 50 | 2/12/2019 | 
| Alice | 100 | 4/23/2019 |
| Mike | 50 | 5/17/2019 |
| Maria | 100 | 6/7/2019 |
| Susanne | 50 | 7/2/2019 |

And this table of donors: 

|Name| Email |
| --- | --- 
| Sam | sam\@gmail.com 
| Alice | aliceemail\@gmail.com
| Susanne | susanneroks\@gmail.com
| Mike | michael\@hotmail.com
| Mary | mary\@gmail.com


We would want to use "Name" to identify which rows correspond to each other in the table. 


Now, how does R decide which order they go in and what gets left in or left out? You need to specify which type of join you want. 


---

## Types of Joins

There are two main types of joins, mutating joins and filtering joins. 

- Mutating joins add all the columns from one table to the other, combining the variables. 

- Filtering joins only take the variables from the first table, but filter the rows based on whether or not that row is present in the second table. 

--- 

### Mutating Joins 

The first kind of join combines variables from both data frames. 


Straight from the R documentation: 

- `left_join(x, y, by = )`: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

- `right_join(x, y, by = )`: return all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

- `inner_join(x, y, by = )`: return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

- `full_join(x, y, by = )`:return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.

How are these different?

---



And again with some simple data. *This data is from the join chapter on stat545.com, linked later in the notes. *

At each step, note what variables and rows appear. 

*Note:* I don't specify which key/identifier variable to use, so R finds variables with the  same names and matches using them. (It prints out what it uses so I can make sure it guessed right.)

```{r}
# make tibbles of data manually

superheroes <- tibble::tribble(
       ~name, ~alignment,  ~gender,          ~publisher,
   "Magneto",      "bad",   "male",            "Marvel",
     "Storm",     "good", "female",            "Marvel",
  "Mystique",      "bad", "female",            "Marvel",
    "Batman",     "good",   "male",                "DC",
     "Joker",      "bad",   "male",                "DC",
  "Catwoman",      "bad", "female",                "DC",
   "Hellboy",     "good",   "male", "Dark Horse Comics"
  )


publishers <- tibble::tribble(
  ~publisher, ~yr_founded,
        "DC",       1934L,
    "Marvel",       1939L,
     "Image",       1992L
  )

superheroes
publishers
```

---

Inner join: Return all rows from x where there are matching values in y, and all columns from x and y.

What is missing? Why? 

```{r, message = F}
inner_join(superheroes, publishers)
```

Think: how will `inner_join(publishers, superheroes)` be different? 

---


```{r}
library(tidyverse)
inner_join(publishers, superheroes)
```

---

Left join: Return all rows from x, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. 

```{r}
left_join(superheroes, publishers)
```

---

Right join: Return all rows from y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. 

*Note:* Not often used. This is the same as `left_join(publishers, superheroes)`

```{r}
right_join(superheroes, publishers)
```

--- 

Full join: Return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing

```{r}
full_join(superheroes, publishers)
```


---

### Filtering Joins 

Filtering joins keep just the cases from the first data frame, but filters using the second. 

- `semi_join(x, y, by = )`: return all rows from x where there are matching values in y, keeping just columns from x.

- `anti_join(x, y, by = )`: return all rows from x where there are not matching values in y, keeping just columns from x.

---

Semi join:  Return all rows from x where there are matching values in y, keeping just columns from x. 

```{r}
semi_join(superheroes, publishers)
```

How is this different than the result from inner_join(superheroes, publishers)?

How will this compare to `semi_join(publishers, superheroes)`?

---

```{r}
semi_join(publishers, superheroes)
```

---

Anti join: Return all rows from x where there are not matching values in y, keeping just columns from x

```{r}
anti_join(superheroes, publishers)
```


I also highly suggest you read through Chapter 11 of our textbook, or Chapter 13 of R for Data Science:  https://r4ds.had.co.nz/relational-data.html
or this lecture from Jenny Bryan: https://stat545.com/join-cheatsheet.html

---

# Examples 

`band_members` and `band_instruments` are two data sets inside R that show these joins pretty well

```{R}
band_members
band_instruments
```

Looking at the two datasets above, convince yourself what each of the different joins below will give you. Once you do so, change `eval = F` to `eval = T`
```{R, eval = F}
inner_join(band_members, band_instruments)
full_join(band_members, band_instruments)
left_join(band_members, band_instruments)
right_join(band_members, band_instruments)

members <- band_members %>% rename(player_name = name)
members
band_instruments
inner_join(members,
           band_instruments,
           join_by(player_name == name))
```


```{r, eval = F}
semi_join(band_members, band_instruments)
anti_join(band_members, band_instruments)

semi_join(band_instruments, band_members)
anti_join(band_instruments, band_members)
```

# Practice

We start by making some "healthcare" data: 

```{r}
# Patients table
patients <- tibble(
  patient_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 40, 35, 50),
  sex = c("F", "M", "F", "M")
)

# Visits table (Some patients have multiple visits, some missing)
visits <- tibble(
  visit_id = c(101, 102, 103, 104, 105),
  patient_id = c(1, 2, 2, 3, 5),  
  doctor_id = c(201, 202, 201, 203, 204),
  visit_date = as.Date(c("2024-01-10", "2024-01-12", "2024-02-15", "2024-03-01", "2024-03-10"))
)

# Doctors table (Some doctors may not have treated any patients)
doctors <- tibble(
  doctor_id = c(201, 202, 203, 205),
  doctor_name = c("Dr. Smith", "Dr. Johnson", "Dr. Lee", "Dr. Brown"),
  specialty = c("Cardiology", "Neurology", "Pediatrics", "Orthopedics")
)

```

Questions: 

Find the names of patients who have visited the hospital. 
```{r}
inner_join(patients,visits)
```

Show all patients visit information, even if they haven't visited the hospital
```{r}
left_join(patients,visits)
```

List all the visits and patient information, even if there isn't information for the patient
```{r}
full_join(patients,visits)
```

Show all visits with their assigned doctors.
```{r}
full_join(visits,doctors)
```

Find the doctor(s) that Bob has met with. 
```{r}
patients %>% 
  filter(name == "Bob") %>%
  full_join(doctors)

left_join(visits,patients) %>% left_join(doctors) %>% filter(name=="Bob") %>% select(name,doctor_name, speciality, everything)
```

Find patients that haven't visited the hospital
```{r}
anti_join(patients,visits)
```

